---
title: "ESM 244 Lab 1"
author: "Sarah Hamilton"
date: "1/5/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf) # install.packages('sf')
library(tmap)
### install.packages("tmap")
### update.packages(ask = FALSE)

# command shift enter to run a whole code chunk
# command enter to run a line
```

### Read in the data

```{r}
sf_trees <- read_csv(here('data','sf_trees','sf_trees.csv'),
                     show_col_types = FALSE) 
```

### Part 1: wrangling and ggplot review

**Example 1:** Find counts of observation by `legal_status` and wrangle a bit

```{r}
### method 1: group_by() %>% summarize()

sf_trees %>%
  group_by(legal_status) %>%
  summarize(tree_count = n())

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>%
  count(legal_status) %>%
  drop_na(legal_status) %>%
  rename(tree_count = n) %>%
  relocate(tree_count) %>%
  slice_max(tree_count, n = 5) %>%
  arrange(-tree_count) 
```

Make a graph of the top 5 from above:

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status,tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen')+
  labs(x = 'legal status', y = 'tree count')+
  coord_flip()+
  theme_minimal()
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA" and store as permitted_data_df

use shift command c to comment or uncomment a line
```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)

permitted_data_df <- sf_trees %>%
  filter(legal_status == 'Permitted Site' & caretaker == 'MTA')
```


**Example 3:** Only keep BLackwood Acacia trees and then only keep columns 'legal_status', 'date','latitude','longitude' and store as `blackwood_acacia_df`

uhhhhhh maybe fix this

```{r}
blackwood_acacia_df <- sf_trees %>%
  filter(str_detect(species, 'Blackwood Acacia')) %>%
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = 'darkgreen')
```

**Example 4:** use `tidyr::separate()`

```{r}
sf_trees_sep <- sf_trees %>%
  separate(species, into = c('spp_scientific','spp_common'),sep = ' :: ')
```

**Example 5:** `ttidyr::unite()`

```{r}
ex_5 <- sf_trees %>%
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```

### Part 2: make some maps

**Step 1:** convert lat and lon to spatial points, `st_as_sf()` function

```{r}
blackwood_acacia_sf <- blackwood_acacia_df %>%
  drop_na(lat,lon) %>%
  st_as_sf(coords = c('lon','lat'))

### we need to tell R what the coordinate reference system is 
st_crs(blackwood_acacia_sf) <- 4326

ggplot(data = blackwood_acacia_sf) +
  geom_sf(color = 'darkgreen')+
  theme_minimal()

```

Read in the SF shapefile and add to map

```{r}
sf_map <- read_sf(here('data','sf_map','tl_2017_06075_roads.shp'))

sf_map_transform <- st_transform(sf_map, 4326)

ggplot(data = sf_map_transform) +
  geom_sf()
```

Continue the maps!

```{r}
ggplot() +
  geom_sf(data = sf_map,
          size = 0.1,
          color = 'darkgray') +
  geom_sf(data = blackwood_acacia_sf,
          color = 'red',
          size = 0.5) +
  theme_void() +
  labs(title = 'Blackwood acacias in SF')
```

### Now an interactive map!

```{r}
tmap_mode('view')

tm_shape(blackwood_acacia_sf) +
  tm_dots()
```


